It’s COMPAS-licated -- Can We Find Fairness in the COMPAS Risk Assessment Algorithm?

### 2022-04-05 Paper_Draft1/Draft2.Rmd

---
title: Does the COMPAS needle always point towards equity? Finding Fairness in the COMPAS Risk Assessment Algorithm
author:
  - name: Amrita Acharya
    affil: 1, 2
  - name: Dianne Caravela
    affil: 1, 3
  - name: Eunice Kim
    affil: 1, 4
  - name: Emma Kornberg
    affil: 1, 5
  - name: Elisabeth Nesmith
    affil: 1, 6
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
  - num: 2
    email: aacharya@smith.edu
  - num: 3
    email: dcaravela@smith.edu
  - num: 4
    email: ekim89@smith.edu
  - num: 5
    email: ekornberg@smith.edu
  - num: 6
    email: enesmith@smith.edu

@@

# Version

This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
However, the official template gets more frequently updated than the 'rticles'
package. Therefore, please make sure prior to paper submission, that you're 
using the most recent .cls, .tex and .bst files 
(available [here](http://www.mdpi.com/authors/latex)).

@@

The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private for-profit company Northpointe, also known by its parent company Equivant [@equivant_faq_nodate], to predict defendants’ risk of recidivism. It generates a score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine]. Jurisdictions across the United States use the COMPAS risk assessment instrument. In 2016, ProPublica published a piece that analyzed the methods and algorithms used by Northpointe to uncover racial biases in defendants’ scores [@angwin2016machine]. Their analysis looks at the distribution of decile COMPAS scores among Black and white defendants. ProPublica concludes, after using a statistical parity metric on false positives or false negatives, that the algorithm is racially biased [@larson2016we]. Northpointe denies the allegations of racial bias, conducting their own analyses based on different statistical parity metrics [@equivant_response_2018]. Since then, ProPublica has responded, maintaining their original argument and Northpointe has yet to respond (though in 2017, they adopted the name Equivant). Currently, the two parties’ disagreement focuses on pretrial Risk Assessment Instruments (RAIs). ProPublica maintains that there are biases in the outcome values, protected attributes, and covariates during Northpointe’s data processing phase. ProPublica accounts for these biases in their analyses, and Northepointe’s response, highlights how ProPublica did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.

@@

# Results

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.

## Subsection Heading Here

Subsection text here.

### Subsubsection Heading Here

Bulleted lists look like this:

* First bullet
* Second bullet
* Third bullet

Numbered lists can be added as follows:

1. First item
2. Second item
3. Third item

The text continues here.

All figures and tables should be cited in the main text as Figure \ref{fig:mdpi-logo}, Table 1, etc.

```{r mdpi-logo, fig.align='center', echo=FALSE, fig.cap="This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\\textbf{a}) Description of what is contained in the first panel. (\\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered."}

# Ben's tutorial on how to include figures with R code chunk
knitr::include_graphics("logo-mdpi.pdf")
```

Please see Table \ref{tab:mytable}.

```{r}
x <- tibble::tribble(~`Title 1`, ~`Title 2`, ~`Title 3`,
"entry 1", "data", "data",
"entry 2", "data", "data"
)

knitr::kable(x, caption = "This is a table caption. Tables should be placed in the main text near to the first time they are cited.\\label{tab:mytable}")
```

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}
<!-- If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph. -->

<!-- Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows: -->
Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

# Discussion

Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

# Conclusion

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

>>>>

---
title: It’s COMPAS-licated -- Can We Find Fairness in the COMPAS Risk Assessment Algorithm?
author:
  - name: Amrita Acharya
    affil: 1, 2
  - name: Dianne Caravela
    affil: 1, 3
  - name: Eunice Kim
    affil: 1, 4
  - name: Emma Kornberg
    affil: 1, 5
  - name: Elisabeth Nesmith
    affil: 1, 6
affiliation:
  - num: 1
    address: |
      Statistical and Data Sciences
      Smith College
      Northampton, MA 01063
  - num: 2
    email: aacharya@smith.edu
  - num: 3
    email: dcaravela@smith.edu
  - num: 4
    email: ekim89@smith.edu
  - num: 5
    email: ekornberg@smith.edu
  - num: 6
    email: enesmith@smith.edu

@@

@@

The Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) algorithm was created by the private for-profit company Northpointe, also known by its parent company Equivant [@equivant_faq_nodate], to predict defendants’ risk of recidivism. It generates a score that classifies defendants’ risk of recidivism as either low, medium, or high [@angwin2016machine]. Jurisdictions across the United States use the COMPAS risk assessment instrument. In 2016, ProPublica published a piece that analyzed the methods and algorithms used by Northpointe to uncover racial biases in defendants’ scores [@angwin2016machine]. Their analysis looks at the distribution of decile COMPAS scores among Black and white defendants. ProPublica concludes, after using a statistical parity metric on false positives or false negatives, that the algorithm is racially biased [@larson2016we]. Northpointe denies the allegations of racial bias, conducting their own analyses based on different statistical parity metrics [@equivant_response_2018]. Since then, the two parties have had several exchanges, maintaining their original arguments. Currently, the two parties’ disagreement focuses on pretrial Risk Assessment Instruments (RAIs). ProPublica maintains that there are biases in the outcome values, protected attributes, and covariates during Northpointe’s data processing phase. ProPublica accounts for these biases in their analyses, and Northepointe’s response, highlights how ProPublica did not account for base rates of recidivism in their analysis, which are important to understand initial percentages without the presence of other information.

@@
